Classless Java

### 2016-03-22 OOPSLA16/sections/Introduction.tex
Intro improved

We present here a third alternative: the concept of \textbf{Interface
  based} (IB) object oriented languages, where objects implement
interfaces directly. In IB interfaces own the implementation for the behaviour, which
is structurally defined in their interface. Programmers do not define objects directly, but
delegate the task to \emph{object interfaces}, whose role is similar
to non abstract classes in CB languages. Key motivations for IB
languages are improved \emph{modularity} and \emph{multiple inheritance}~\bruno{I'm using multiple
  inheritance in a more abstract way than you Marco.}\marco{Yes, but will the reviewer think in the abstract or in the
  concrete way about MI?}  The literature
provides good examples on how easy and modular it is to combine
multiple sources of pure behaviour, using mechanisms such as
traits~\cite{scharli03traits}. 
In Java multiple \emph{interface} inheritance has been
supported since inception, and in Java8 default methods~\cite{} bring some of
the advantages of traits into Java. In contrast, the literature~\cite{} is
also rich on how hard it is to modularly combine multiple sources of
behaviour \textbf{and} state with multiple \emph{implementation}
inheritance of classes. 
%Traits or Java8 interfaces still assume a CB model: 
%But traits still require classes, which
%are responsible for object construction and adding state.

To retain the easiness and modularity of combining multiple sources of
pure behaviour, in IB state is just a special kind of behaviour
implementation.
\marco{NO NO!
the advantage is not just multiple composition 
(by the way, that is the term you are searching for, not multiple inheritance)
By not having state reified in the ideal/platonic level, it is not fixed
what is state and what is not, thus you can (model an encoding of) REMOVE fields and you can refine
fields.}
 Objects are the only responsibles to define the
ultimate behaviour of a method and, for example, if such method is
just a setter. Anything related to state is completely contained in
the instances and do not leak in the interfaces.\bruno{Not sure if
  this is totally clear. Implicit here is the fact that objects are
  never defined explicitly, I think. And the behaviour is introduced 
implicitly.}
\marco{we need to discuss, not clear what you mean}
%\footnote{ In the same sense,
%  you can encode prototype based programming over class based with the
%  strategy pattern, and class based programming with prototypes by
%  creating ``class'' objects.}

@@

In the context of Java, what we propose is essentially a programming
style, where we never use classes\footnote{More precisely, we never
  use the \Q@class@ keyword.}.  We call this restricted version of
Java \emph{Classless Java}.  
Our proposed style enables powerful
object-oriented idioms, similar to (stateful) multiple inheritance and
field type refinement (allowing a kind of covariant setters
refinement). \bruno{needs strenghtening}

A key challenge lies in how to model state, which is fundamental to
have stateful objects. In IB and Classless Java, the idea is that all abstract operations in an Object
interface will be interpreted as \emph{abstract state operations}. The
abstract state operations include various common utility methods (such
as getters and setters, or clone-like methods). In the presence of
subtyping, such operations often require special care, as their types
need to be refined. Object interfaces provide support for
type-refinement and can automatically produce code that deals with
type-refinement adequately, while Java (and Scala) require explicit
type-refinement.\bruno{strenghtening needed!}

>>>>

We present here a third alternative: the concept of \textbf{Interface
  based} (IB) object oriented languages, where objects implement
interfaces directly. In IB interfaces own the implementation for the behaviour, which
is structurally defined in their interface.

\marcoT{
A key challenge lies in how to model state, which is fundamental to
have stateful objects. 
Programmers do not define objects directly, but
delegate the task to \emph{object interfaces}, whose role is similar
to non abstract classes in CB languages.
All abstract operations in an object interface will be interpreted as \emph{abstract state operations}. The
abstract state operations include various common utility methods (such
as getters and setters, or clone-like methods). In the presence of
subtyping, such operations often require special care, as their types
need to be refined. Object interfaces provide support for
type-refinement and can automatically produce code that deals with
type-refinement adequately, while CB (as in Java, Scala and C\#) require 
verbose explicit type-refinement.
We believe that such verbosity hindered and slowed down
the discovery of useful programming pattens involving type-refinement\cite{ trivial solution to the EP}.
}

Finally, object interfaces also provide static factory methods instantiating objects.

\marcoT{%
Objects are the only responsible to define the
ultimate behaviour of a method and, for example, if such method is
just a setter. Anything related to state is completely contained in
the instances and do not leak in the inheritance logic.
In CB, the structure of the state is fixed and can be only
grown by inheritance; on the opposite, in IB the state is never fixed,
and methods that ``looks like'' abstract setters and getters
can always receive an explicit implementation down in the inheritance chain,
improving \textbf{modularity and flexibility}.

That is, the concept of abstract state is more fluid.
In this paper we show how to support type-safe
\textbf{field removal},
\textbf{field type refinement} allowing a kind of covariant setters refinement,
and \textbf{multiple inheritance}.}

Indeed the literature
provides good examples on how easy and modular it is to combine
multiple sources of pure behaviour, using mechanisms such as
traits~\cite{scharli03traits}. 
In Java multiple \emph{interface} inheritance has been
supported since inception, and in Java8 default methods~\cite{} bring some of
the advantages of traits into Java. In contrast, the literature~\cite{} is
also rich on how hard it is to modularly combine multiple sources of
behaviour \textbf{and} state with multiple \emph{implementation}
inheritance of classes. 
%Traits or Java8 interfaces still assume a CB model: 
%But traits still require classes, which
%are responsible for object construction and adding state.

%To retain the easiness and modularity of combining multiple sources of
%pure behaviour, in IB state is just a special kind of behaviour
%implementation.
%\marco{NO NO!
%the advantage is not just multiple composition 
%(by the way, that is the term you are searching for, not multiple inheritance)
%By not having state reified in the ideal/platonic level, it is not fixed
%what is state and what is not, thus you can (model an encoding of) REMOVE fields and you can refine
%fields.}
% Objects are the only responsibles to define the
%ultimate behaviour of a method and, for example, if such method is
%just a setter. Anything related to state is completely contained in
%the instances and do not leak in the interfaces.

\bruno{Not sure if
  this is totally clear. Implicit here is the fact that objects are
  never defined explicitly, I think. And the behaviour is introduced 
implicitly.}
\marco{we need to discuss, not clear what you mean}
%\footnote{ In the same sense,
%  you can encode prototype based programming over class based with the
%  strategy pattern, and class based programming with prototypes by
%  creating ``class'' objects.}

@@

In the context of Java, what we propose is essentially a programming
style, where we never use classes\footnote{More precisely, we never
  use the \Q@class@ keyword.}.  We call this restricted version of
Java \emph{Classless Java}.
